Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-08-29 src/tex/ms.tex
edits

\citet{Wang_2021} performed \ac{PE} on GWTC-1 events with a frequency-dependent model of birefringence resembling ours.
Following \cite{Zhao:2019xmm}, that reference parameterized the birefringent waveform in terms of some amplitude and phase modifications to the linear polarizations ($\delta h$ and $\delta \Psi$ respectively), such that
\begin{equation}
    h_{+/\times}^{\rm BR}(f) = h_{+/\times}^{\rm GR}(f)\mp h_{\times/+}^{\rm GR}(f)(i\delta h-\delta\Psi)\,,
\end{equation}
with the plus (minus) sign for the cross (plus) polarization.
In terms of circular polarizations, this is equivalent to 
\begin{align}
h^{\rm BR}_{L/R}(f) &= h^{\rm GR}_{L/R}(f) \left(1 \mp \delta h \mp i \delta \Psi \right) \nonumber\\
&\approx h^{\rm GR}_{L/R}(f) \exp(\mp \delta h \mp i \delta \Psi) \, ,
\end{align}
assuming a small $\delta h$ and $\delta \Psi$ in the last line.
To consider only amplitude birefringence, as in this work, we must compare to the result in Ref.~\citep{Wang_2021} that set $\delta \Psi = 0$ and parameterized $\delta h = \pi f z / M_{\rm PV}$, where $M_{\rm PV}$ is the energy scale of the birefringent (parity-violating) correction.%
\footnote{Concretely, \citet{Wang_2021} write $\delta h = - A_\nu \pi f$ with $A_\nu = M_{\rm PV}^{-1} \left[\alpha_\nu (z=0) - \alpha_\nu(z) \left(1+z\right)\right]$, for $\alpha_\nu$ some function of redshift encoding the evolution of a birefringence-mediating field; to produce their constraint, however, they further set $\alpha_\nu(z) = 1$, yielding $A_\nu = -z / M_{\rm PV}$ and hence $\delta h = \pi f z / M_{\rm PV}$.}
Comparing to our Eq.~\eqref{eq:waveform_modification} and approximating $z \approx d_C H_0/c$ via the Hubble constant, $H_0$, this means
\begin{equation}
    |\kappa| \simeq \frac{\pi H_0}{c} \frac{h \left( 100 \, \mathrm{Hz} \right)}{M_{\rm PV}} \simeq \frac{3.0 \times 10^{-22} \, \mathrm{GeV}}{M_\mathrm{PV}} \, \mathrm{Gpc}^{-1},
\end{equation}
where $h$ is Planck's constant.
\citet{Wang_2021} quote a constraint of $M_{\rm PV} > 10^{-22}\, {\rm GeV}$ at 90\% credibility, which translates into \variable{output/kappa_Wang.txt}.
We obtained a tighter 90\% upper limit of \variable{output/restricted_absolute_kappa_90.txt}, or \variable{output/M_PV_constraint.txt}.
Note that our constraints from each event are stronger compared to the constraint from \citet{Wang_2021}, details of which are given in Sec.~\ref{sec:comparison_Zhu}.

@@

\subsubsection{\added{Zhu et al.}}
\label{sec:comparison_Zhu}

A more recent work by \citet{Zhu:2023wci}, which appeared while this manuscript was finalized, performed \ac{PE} on GWTC-3 events with a frequency-dependent model of amplitude birefringence.
Their model is the same as used in \citet{Wang_2021}, which allows us to compare our results, as we did in Sec.~\ref{sec:comparison_Wang}.
\citet{Zhu:2023wci} reported a constraint of $M_{\rm PV} > 4.1 \times 10^{-22}\, {\rm GeV}$ at 90\% credibility, which translates into \variable{output/kappa_Zhu.txt}.
This result constrains the birefringence effect much weaker than our result.
This is an unexpected result, as the analysis in \citet{Zhu:2023wci} is similar to ours\footnote{Note that the prior on the birefringence parameter and luminosity distance are different between the analysis by \citet{Zhu:2023wci}and ours, but it is unlikely to be the cause of the difference in the constraint.} and they included more events than we did.

In \ref{fig:posterior_MPV}, we show the posterior on $M_{\rm PV}^{-1}$ for each event in our analysis.
Compared to Fig.~1 in \citet{Zhu:2023wci}, we can see that the posteriors on some events are different between our analysis and theirs.
For example, the posterior for GW190727\_060333 peaks at zero in our analysis, but peaks at a nonzero value in \citet{Zhu:2023wci}.
This shows that the difference in the combined constraint may come from the difference in the individual event analyses.
However, we could not compare the analysis in \citet{Zhu:2023wci} with ours in detail, as their analysis result is not publicly available.

\begin{figure}
    \script{posterior_MPV.py}
    \includegraphics[width=\columnwidth]{figures/posterior_MPV.pdf}
    \caption{
        Individual-event $M_\mathrm{PV}^{-1}$ posteriors from our analysis, converted from the $\kappa$ posteriors in Fig.~\ref{fig:violin_kappa}.
        Chosen events are those labeled in Fig.~1 of \citet{Zhu:2023wci}.
    }
    \label{fig:posterior_MPV}
\end{figure}

>>>>

\citet{Wang_2021} performed \ac{PE} on GWTC-1 events with a frequency-dependent model of birefringence resembling ours.
Following \cite{Zhao:2019xmm}, that reference parameterized the birefringent waveform in terms of some amplitude and phase modifications to the linear polarizations ($\delta h$ and $\delta \Psi$ respectively), such that
\begin{equation}
    h_{+/\times}^{\rm BR}(f) = h_{+/\times}^{\rm GR}(f)\mp h_{\times/+}^{\rm GR}(f)(i\delta h-\delta\Psi)\,,
\end{equation}
with the plus (minus) sign for the cross (plus) polarization.
In terms of circular polarizations, this is equivalent to 
\begin{align}
h^{\rm BR}_{L/R}(f) &= h^{\rm GR}_{L/R}(f) \left(1 \mp \delta h \mp i \delta \Psi \right) \nonumber\\
&\approx h^{\rm GR}_{L/R}(f) \exp(\mp \delta h \mp i \delta \Psi) \, ,
\end{align}
assuming a small $\delta h$ and $\delta \Psi$ in the last line.
To consider only amplitude birefringence, as in this work, we must compare to the result in Ref.~\citep{Wang_2021} that set $\delta \Psi = 0$ and parameterized $\delta h = \pi f z / M_{\rm PV}$, where $M_{\rm PV}$ is the energy scale of the birefringent (parity-violating) correction.%
\footnote{Concretely, \citet{Wang_2021} write $\delta h = - A_\nu \pi f$ with $A_\nu = M_{\rm PV}^{-1} \left[\alpha_\nu (z=0) - \alpha_\nu(z) \left(1+z\right)\right]$, for $\alpha_\nu$ some function of redshift encoding the evolution of a birefringence-mediating field; to produce their constraint, however, they further set $\alpha_\nu(z) = 1$, yielding $A_\nu = -z / M_{\rm PV}$ and hence $\delta h = \pi f z / M_{\rm PV}$.}
Comparing to our Eq.~\eqref{eq:waveform_modification} and approximating $z \approx d_C H_0/c$ via the Hubble constant, $H_0$, this means
\begin{equation} \label{eq:kappa-to-Mpv}
    |\kappa| \simeq \frac{\pi H_0}{c} \frac{h \left( 100 \, \mathrm{Hz} \right)}{M_{\rm PV}} \simeq \frac{3.0 \times 10^{-22} \, \mathrm{GeV}}{M_\mathrm{PV}} \, \mathrm{Gpc}^{-1},
\end{equation}
where $h$ is Planck's constant.
\citet{Wang_2021} quote a constraint of $M_{\rm PV} > 10^{-22}\, {\rm GeV}$ at 90\% credibility, which translates into \variable{output/kappa_Wang.txt}.
We obtained a tighter 90\% upper limit of \variable{output/restricted_absolute_kappa_90.txt}, or \variable{output/M_PV_constraint.txt}.

@@

\subsubsection{\added{Zhu et al.}}

A more recent work by \citet{Zhu:2023wci}, which appeared while this manuscript was finalized, performed \ac{PE} on GWTC-3 events with a frequency-dependent model of amplitude birefringence.
Their model is the same as the one used in \citet{Wang_2021}, which allows us to compare to our results as we did in Sec.~\ref{sec:comparison_Wang}.
\citet{Zhu:2023wci} reported a constraint of $M_{\rm PV} > 4.1 \times 10^{-22}\, {\rm GeV}$ at 90\% credibility, which translates into \variable{output/kappa_Zhu.txt} per Eq.~\eqref{eq:kappa-to-Mpv} above.

This constraint is weaker than ours in spite of the similar number of events considered, a discrepancy that might be explained by a number of factors.
First, there are difference in priors, since \citet{Zhu:2023wci} assume a distribution uniform in $z/M_{\rm PV}$ rather than $\kappa$ or $M_{\rm PV}$ itself.
Second, as reported in \citep{Zhu:2023wci}, their combined result was computed by multiplying $M_{\rm PV}^{-1}$ posteriors from individual events; however, this is not the correct way of combining $M_{\rm PV}$ measurements from independent observations: instead of multiplying posteriors, one should multiply the \emph{likelihoods}Z to avoid double counting the prior every time a new event is added to the combined constraint.
Finally, \citet{Zhu:2023wci} report \emph{ad hoc} exclusion of certain events from the combined constraint, without explicitly providing a list of events considered.
All these factors combined preclude a clear interpretation of the results in \citep{Zhu:2023wci}.
